### 2021-12-05 documents/sections/proto_carib.tex
data output
rewriting and restructuring

\section{Cariban verbal person marking and marker extensions}
The archaic first person markers investigated in this paper are the result of incomplete person marker extensions, i.e., modifications to the ancestral prefix system that did not affect the whole lexicon.
To provide readers with the necessary background, we first introduce said ancestral system \pcref{sec:pc_person}, then the peculiarities of the Cariban split-\gl{s} system \pcref{sec:split}, and finally the concept of person marker extensions \pcref{sec:extensions_intro}.

\subsection{Verbal person marking in \PC}
\label{sec:pc_person}
\PC is reconstructed by \textcite{gildea1998} as using a person paradigm called \setone in its independent verb forms, shown in \cref{tab:pcpers}.%
\footnote{We use standard IPA symbols in our transcription of Cariban languages, with the exception of coronal rhotics, which we simply represent with \ort{r}, rather than \ort{ɽ} for \wayana or \ort{ɾ̠} for \maqui etc.
In languages with strong morphophonological processes and/or subphonemic orthography we show the original transcription in an additional surface line when presented in an interlinearized glossed example.
We follow \textcite{gildea2018reconstructing} in using \ort{ə} for the proto-vowel reconstructed by \textcite{meira2005southern}, although it was likely more back \parencite{gildea2010story}.}
The choice of person marker in transitive verbs can be characterized as being conditioned by a basic person hierarchy \fbox{\gl{1}/\gl{2} > \gl{3}}.\footnote{Glossing abbreviations: \glossingAbbrevsComma.}
The locuphoric markers had two forms, an \gl{a}-oriented one for direct (\gl{sap}>\gl{3}) scenarios and a \gl{p}-oriented one for inverse (\gl{3}>\gl{sap}) scenarios.
There was a single aliophoric marker \rc{n(i)-}, which only surfaced in nonlocal (\gl{3}>\gl{3}) scenarios, without morphologically expressed distinctions between different third person referents.
Local scenarios were expressed in a non-transparent manner, by using the \gl{1+2} prefix \rc{k-} in both cases.%
\footnote{The presence of a \gl{1+2} person value implies that of a \gl{1+3} value.
This is usually expressed with a free pronoun combined with third person morphology in Cariban languages, so it is not represented in this table.}

\begin{table}
	\centering
	\caption[\PC \setone (main clause) person markers]{\PC \setone (main clause) person markers \parencites[495]{meira2010origin}[497]{gildea2016referential}}
	\label{tab:pcpers}
\begin{subtable}[b]{.49\linewidth}
\caption{Transitive}
\label{tab:pctrans}
\centering
	\begin{tabular}{@{}lllll@{}}
	\mytoprule
\gl{a}/\gl{p}&		\gl{1}	&	\gl{2}		&	\gl{1+2}	&	\gl{3}	\\
\mymidrule
\gl{1}	&		&	\rc{k-}	&				&	\rc{t(i)-}		\\	
\gl{2}	&	\rc{k-}			&&				&	\rc{m(i)-}		\\
\gl{1+2}&		&				&				&	\rc{kɨt(i)-}		\\
\gl{3}	&	\rc{u(j)-}	&	\rc{ə(j)-}	&	\rc{k-}			&	\rc{n(i)-}		\\
	\mybottomrule
	\end{tabular}
\end{subtable}%
\begin{subtable}[b]{.49\linewidth}
\caption{Intransitive}
\label{tab:pcintrans}
\centering
\begin{tabular}{@{}lll@{}}
\mytoprule
& \gl{s_a_} & \gl{s_p_}  \\
\mymidrule
\gl{1} & \rc{w-} & \rc{u(j)-} \\
\gl{2} & \rc{m-} & \rc{ə(j)-}\\
\gl{1+2} & \rc{kɨt-} & \rc{k-}\\
\gl{3} & \rc{n-} & \rc{n(i)-}\\
\mybottomrule
\end{tabular}	
\end{subtable}
\end{table}

@@

As in many other split-\gl{s} systems found in languages of the world, the intransitive verbal lexicon was divided into two classes, with verbs inherently being members of the \gl{s_a_} or the \gl{s_p_} group.
That is, while in transitive verbs, the choice of person marker had a crucial semantic contribution, it was predictable in intransitive verbs.
This is illustrated with modern \kalina data in \exref[kar1]{kar2}.

\pex<kar1> \kalina
\a<kar-60>
\begingl
\gla mi-kupi-ja//
\glb \gl{2}>\gl{3}-bathe-\gl{prs}//
\glft \qu{You bathe him/her.} \parencite[][160]{hoff1968carib}//
\endgl
\a<kar-62>
\begingl
\gla a-kupi-ja//
\glb \gl{3}>\gl{2}-bathe-\gl{prs}//
\glft \qu{S/he bathes you.} \parencite[][63]{yamada2011evidentiality}//
\endgl
\xe
%
In \exref{kar1}, the choice between the second person \gl{a}- and \gl{p}-oriented markers \obj{mi-} and \obj{a-} depends on the scenario:
The transitive verb \obj{kupi} \qu{to bathe} takes \obj{mi-} in \gl{2}>\gl{3} scenarios \exref{kar1.kar-60}, but \obj{a-} in \gl{3}>\gl{2} scenarios \exref{kar1.kar-62}.
While intransitive verbs show the same (or very similar) person markers, they contribute no semantic difference here \exref{kar2}.

\pex<kar2> \kalina
\a<kar-59>
\begingl
\gla sipi tɨnka-rɨ m-ekema-non hen//
\glb net pull-\gl{nmlz} \gl{2}-be.afraid-\gl{prs}.\gl{uncert} eh?//
\glft \qu{You're afraid to pull up the net, aren't you? } \parencite[][253]{courtz2008carib}//
\endgl
\a<kar-61>
\begingl
\gla aj-awoi-ja//
\glb \gl{2}-get.up-\gl{prs}//
\glft \qu{You are getting up.} \parencite[][167]{hoff1968carib}//
\endgl
\xe
%
Rather, \obj{ekema} \qu{to be afraid} takes an \gl{a}-oriented marker, since it is an \gl{s_a_} verb \exref{kar2.kar-59}, while the \gl{s_p_} verb \obj{awomɨ} \qu{to get up} takes a \gl{p}-oriented marker \exref{kar2.kar-61}.\footnote{The root \obj{awomɨ} \qu{to get up} is subject to syllable reduction and assimilation to the prefix-initial \obj{j}.}
What is more, the choice of prefix does not necessarily convey information about the semantics of the verb, since there are clear mismatches:
\qu{to be afraid} with an \dbqu{agentive} marker can hardly be considered a volitional act, while  \qu{to get up} with a \dbqu{patientive} marker must be considered volitional.
These mismatches are not isolated cases, as will be discussed in \cref{sec:split}.

\subsection{Defining features and origins of the split-\gl{s} system}
\label{sec:split}
As seen in the previous section, the split-\gl{s} system defined two inflectional classes for intransitive verbs within the \setone system.
%\kalina paradigms illustrating the split for each person are given in \cref{tab:kalinaparadigms}.
%\gl{3}\gl{s_a_} differs from \gl{3}\gl{s_p_} by \begin{inlinelist}
%\item not showing the allomorph \obj{ni-} before consonants
%\item not triggering the umlaut of verb-initial \phon{\obj{o}}{\obj{e}}
%\end{inlinelist}, which is shown in \cref{tab:kalinaparadigms}.
%
%\begin{table}[h]
%	\centering
%	\caption{\kalina \gl{s_a_} and \gl{s_p_} person paradigms \parencites[178]{hoff1968carib}[430, 73]{courtz2008carib}[481]{meira2010origin}}
%	\label{tab:kalinaparadigms}
%	\begin{tabular}{@{}lll@{}}
%	\mytoprule
%Person & \gl{s_a_} \obj{opɨ} \qu{to come} & \gl{s_p_} \obj{[o/e]kanumɨ} \qu{to run}\\
%	\mymidrule
%\gl{1} & \obj{w-opɨ-} & \obj{j-ekanumɨ-}\\
%\gl{2} & \obj{m-opɨ-} & \obj{aj-ekanumɨ-}\\
%\gl{1+2} & \obj{kɨt-opɨ-} & \obj{k-okanumɨ-}\\
%\gl{3} & \obj{n-opɨ-} & \obj{n-ekajumɨ-}\\
%	\mybottomrule
%	\end{tabular}
%\end{table}
%
However, there were some other morphological criteria distinguishing \gl{s_a_} from \gl{s_p_} verbs in \PC:
\begin{inlinelist}
\item presence vs absence of the \gl{s_a_} marker \rc{w-}
\item absence vs presence of the second person prefix \rc{ə(j)-} in imperatives
\item presence vs absence of a derivational detransitivizing prefix
\end{inlinelist}.

Many languages show an \gl{s_a_} class marker in deverbalized forms, which can be reconstructed to \PC as \rc{w-}.\footnote{See \textcite[227]{meira2000split}, who identifies reflexes of this morpheme as having \dbqu{no purpose other than being \qu{class markers}, without any obvious semantic or functional load}.}
With \gl{s_a_} verbs, \rc{w-} occurred immediately between the possessive prefixes and the verb stem, while \gl{s_p_} verbs took the bare prefixes.
Reflexes of \rc{w-} in languages from different branches are illustrated in \cref{tab:participles,tab:nominalizations} for participles and nominalizations.
%\kaxui does not preserve \rc{w-} in its participles (\obj{tehurkat͡ʃe} \qu{fallen}, not \ungr{tɨwehurkat͡ʃe}), while \apalai only preserves it there, but not in nominalizations (\obj{jepɨtopo} \qu{my bathing place}, not \ungr{joepɨtopo}).

\input{floats/participles}

\input{floats/nominalizations}

The distinction between \gl{s_a_} and \gl{s_p_} is also borne out in imperatives.
Here, \gl{s_p_} verbs took the \gl{p}-oriented second person prefix \rc{ə(j)-}, while \gl{s_a_} verbs were unprefixed (both suffixed with \rc{-kə}).
This is illustrated with reflexes in various modern languages in \cref{tab:imperatives}.
As in the case of the \gl{s_a_} marker \rc{w-} participles and nominalizations, some languages have lost the distinction between \gl{s_a_} and \gl{s_p_} verbs in imperatives, for example \panare.

\input{floats/imperatives}

%\subsubsection{The detransitivizing prefix}
There is one further property uniting most \gl{s_a_} verbs, not based on inflectional morphology.
As mentioned in \cref{sec:pc_person}, mismatches between the semantics of intransitive verbs and their \gl{a}- or \gl{p}-oriented inflectional morphology are common.
However, the Cariban split-\gl{s} system goes further than all other known such systems, in that its division of the verbal lexicon does not follow any discernible semantic criteria whatsoever.
\textcite{meira2000split} takes a sizable corpus of intransitive verbs from \trio, \kalina, \apalai, and \wayana, and categorizes them by applying different criteria commonly encountered in split-\gl{s} systems.
He shows that neither (non\-)activities, % \parencite[210]{meira2000split}, 
(non-)agency, % \citeyearpar[212]{meira2000split}, 
(in-)animacy, % \citeyearpar[213]{meira2000split},
nor Aktionsart % \citeyearpar[215]{meira2000split} 
satisfactorily predict the class membership of intransitive verbs.

Rather, the reason for a verb to take the \gl{a}- or \gl{p}-oriented prefix is (at least diachronically) a morphological one.
\textcite[217--221]{meira2000split} demonstrates that those intransitive verbs which (etymologically) have a detransitivizing prefix are treated as \gl{s_a_} verbs, while essentially all others are \gl{s_p_} verbs:
\begin{quotebox}{\parencite[201]{meira2000split}}
Almost all verbs in the \gl{s_a_} class are detransitivized forms of transitive verbs, either synchronically (with still exisiting transitive sources) or diachronically (with reconstructible but no longer existing transitive sources)
\end{quotebox}
\textcite[221--223]{meira2000split} also argues that the detransitivizing prefixes are indeed deriving \gl{s_a_} verbs, rather than being inflectional in nature:
\begin{inlinelist}
	\item there are a few underived \gl{s_a_} verbs, with no detransitivizing prefix
	\item \gl{s_a_} verbs can develop irregular semantics compared to their transitive counterparts
	\item it is unpredictable whether the \gl{a} or \gl{p} argument of the underlying transitive verb becomes the \gl{s} of the derived \gl{s_a_} verb
	\item some originally derived \gl{s_a_} verbs have lost their transitive counterparts
	\item \dbqu{basic} concepts are expressed as derivations of more complex concepts, like \qu{to dance (\gl{s_a_})} from \qu{to dance with (\gl{tr})}
\end{inlinelist}.
He also notes that this leads to an inflectional split not based in meaning, but rather morphology:

@@

As for the form of this marker, \textcite[505--512]{meira2010origin} reconstruct two distinct  prefixes for \PC: reciprocal \rc{əte-} and reflexive \rc{e-}, although they have since merged into a single morpheme, apparently in all languages.
Modern reflexes of \detrz show a range of meanings, which can all be characterized as \dbqu{detransitive}; this range is illustrated with \trio examples in \exref{triodetrz}.

\ex<triodetrz> \trio \parencites[218--219]{meira2000split}[128, 256]{triomeira1999}\\
\begin{tabular}[t]{@{}llll@{}}
\\
\obj{nonta}  & → & \obj{e-nonta}, & \qu{abandon each other}\\
\qu{abandon} & & \obj{əi-nonta} &  (reciprocal) \\
\\
\obj{suka} & → & \obj{e-suka}, & \qu{wash self}\\
\qu{wash} & & \obj{əi-suka} & (reflexive)\\
\\
\obj{pahka} & → & \obj{e-pahka} & \qu{break (\gl{intr})}\\
\qu{break (\gl{tr})} & & & (anticausative)\\
\\
\obj{puunəpɨ} & → & \obj{əh-puunəpɨ}, & \qu{think, meditate}\\
\qu{think about} & & \obj{əi-puunəpɨ} & (antipassive)\\
\end{tabular}
\xe
%
The morphological variation featured in \qu{to abandon each other} and \qu{to wash self} is due to the mentioned collapse between the two \PC prefixes:
\obj{e-} is a reflex of the reflexive prefix \rc{e-}, while the form \obj{əi-} originates in reciprocal \rc{əte-}.
However, both can occur with either meaning -- at least for these two verbs.

%\subsubsection{Underived \gl{s_a_} verbs}
As mentioned, there are a few \gl{s_a_} verbs that are not derived from transitive verbs, and do not have a reflex of \detrz, either.
\textcite[221]{meira2000split} characterizes these as very old and exceptional, and counts between 5 and 7 verbs, depending on the language.
\textcite{gildea2007greenberg} identify 7 such underived \gl{s_a_} verbs as reconstructible to \PC, shown in \cref{tab:underived2007}.
Many of these verbs will be discussed in \cref{sec:verbs}.

\begin{table}
	\centering
	\caption[Underived \PC \gl{s_a_} verbs]{Underived \PC \gl{s_a_} verbs \parencite[30]{gildea2007greenberg}}
	\label{tab:underived2007}
	\begin{tabular}{@{}lll@{}}
	\mytoprule
Form & Meaning\\
\mymidrule
\rc{tə[mə]} & \qu{to go}\\
\rc{ətepɨ} & \qu{to come\subs{1}}\\
\rc{ka[ti]} & \qu{to say}\\
\rc{əmə[mɨ]} & \qu{to enter}\\
\rc{eti} & \qu{to dwell, be\subs{2}}\\
\rc{a[p]} & \qu{to be\subs{1}, say}\\
\rc{əməkɨ} & \qu{to come\subs{2}}\\
	\mybottomrule
	\end{tabular}
\end{table}

\subsection{Person marker extensions in intransitive verbs}
\label{sec:extensions_intro}
The original \PC split-\gl{s} system has been subject to change in many languages, in some even to the point of total loss.
Besides modifications like losing the \gl{s_a_} class marker \rc{w-} or losing certain person values altogether, these changes have largely been due to person prefixes being extended to new verbs.\footnote{
In principle, one could also imagine a scenario in which verbs of one of the classes are gradually replaced with innovative verbs from the other class, as suggested for \panare by \textcite[225]{meira2000split}.
However, the verbs Meira sampled came from the \emp{A}-section of \posscite{mattei1994diccionario} dictionary \perscommpar{Spike Gildea}, and \obj{at-}\slash{}\obj{at͡ʃ-}\slash{}\obj{as-} is a frequent reflex of \detrz in \panare, meaning that many \gl{s_a_} verbs are \obj{a}-initial, thus explaining the 80:40 \gl{s_a_}:\gl{s_p_} ratio in Meira's sample.
Thus, the hypothesis that \panare has largely replaced \gl{s_p_} with \gl{s_a_} verbs remains to be thoroughly tested.
For the moment, the most likely scenario leading to the eventual loss of the split-\gl{s} system involves markers being extended to new verbs.}
There have been many such person marker extensions in Cariban languages, and some are still ongoing.
This is shown by \textcite{gildea1998}, using the three Parukotoan languages as an example.
We have reproduced his tables as a tree diagram in \cref{fig:par_ext}, with adapted transcription -- and in the case of \kaxui, the addition of ∅/\obj{j-} as an alternative \gl{1}\gl{s_p_} marker \perscommpar{Spike Gildea}.
Apart from segmental changes to individual morphemes, the following restructuring innovations happened in the \setone paradigm in Parukotoan:

>>>>

\section{The origins of conservative first person inflections}
\label{sec:background}
The irregular first person prefixes introduced in \cref{sec:intro} are relics, inherited from the ancestral \PC system \pcref{sec:pc_person}.
That system was subject to many different kinds of innovations; responsible for the patterns discussed are person marker extensions not spreading through the entire \gl{s_a_} lexicon \pcref{sec:extensions_intro}.
A particular component of the system, the distinction between \gl{s_a_} and \gl{s_p_} verbs, plays a role in what extensions are incomplete and is the topic of \cref{sec:split}.
%\cref{sec:outlook} summarizes the main points of this section and provides an outlook for the rest of the paper.

\subsection{\PC person marking and inflectional relics}
\label{sec:pc_person}
\PC is reconstructed by \textcite{gildea1998} as using a person paradigm called \setone in its independent verb forms, shown in \cref{tab:pcpers}.
The choice of person marker in transitive verbs can be characterized as being conditioned by a basic person hierarchy \fbox{\gl{1}/\gl{2} > \gl{3}}.
The locuphoric markers had two forms, an \gl{a}-oriented one for direct (\gl{sap}>\gl{3}) scenarios and a \gl{p}-oriented one for inverse (\gl{3}>\gl{sap}) scenarios.
There was a single aliophoric marker \rc{n(i)-}, which only surfaced in nonlocal (\gl{3}>\gl{3}) scenarios, without morphologically expressed distinctions between different third person referents.
Local scenarios were expressed in a non-transparent manner, both using the \gl{1+2} prefix \rc{k-}.

\input{floats/pcpers}

@@

Equipped with this knowledge about the ancestral system, it becomes clear that the divergent \hixka and \trio forms in \cref{tab:hixintro,tab:triintro} behave irregularly because they preserve the original \PC \gl{1}\gl{s_a_} prefix \rc{w-}; they are therefore \textsc{conservative}.
They contrast with regular \gl{s_a_} verbs, which are innovative in both languages.
The reflexes of \rc{w-} may be considered \textsc{relics}, old and restricted to specific lexically conditioned contexts, contrasting with the innovative prefixes found elsewhere.
These verbs and their prefixes are comparable with the few English nouns like \obj{ɒks}, which preserve the old plural suffix \obj{-ən}.
It, too, was once more widespread, being the normal plural suffix of the weak inflection, compare German \obj{ɔks-n} \qu{ox-en}, \obj{naːmə-n} \qu{name-s}, \obj{haːzə-n} \qu{hare-s}, \obj{bɛːʁ-ən} \qu{bear-s}.
Since the irregular \hixka and \trio prefixes are conservative and the regular prefixes are innovative, the next question to be addressed is where these new prefixes came from.

%\input{floats/makintro}

\subsection{Person marker extensions in intransitive verbs}
\label{sec:extensions_intro}
In his discussion of the \PC split-\gl{s} system and reconstruction of the intransitive person prefixes, \textcite[88--96]{gildea1998} shows that the system has undergone many different modifications in various languages.
The main mechanism of change leading to these modifications are \textbf{person marker extensions}, i.e. the use of verbal person prefixes being extended to contexts previously occupied by other prefixes.
%\footnote{
%In principle, one could also imagine a scenario in which verbs of one of the classes are gradually replaced with innovative verbs from the other class, as suggested for \panare by \textcite[225]{meira2000split}.
%However, the verbs Meira sampled came from the \emp{A}-section of \posscite{mattei1994diccionario} dictionary \perscommpar{Spike Gildea}, and \obj{at-}\slash{}\obj{at͡ʃ-}\slash{}\obj{as-} is a frequent reflex of \detrz in \panare, meaning that many \gl{s_a_} verbs are \obj{a}-initial, thus explaining the 80:40 \gl{s_a_}:\gl{s_p_} ratio in Meira's sample.
%Thus, the hypothesis that \panare has largely replaced \gl{s_p_} with \gl{s_a_} verbs remains to be thoroughly tested.
%For the moment, the most likely scenario leading to the eventual loss of the split-\gl{s} system involves markers being extended to new verbs.}
There have been many different person marker extensions in Cariban languages, and some are still ongoing.
This is illustrated by \textcite{gildea1998}, using the three Parukotoan languages as an example.
Apart from segmental changes to individual morphemes, the following innovations happened in the \setone paradigm in Parukotoan:

\begin{enumerate}
	\item \PPar \begin{enumerate}
		\item \gl{1}\gl{s_a_} \rc{w-} to \gl{1}>\gl{3}
		\item \gl{1+2} \rc{k-} to \gl{1}\gl{s_p_} (completed in \PWai, ongoing in \kaxui)
		\item \gl{1+2} \rc{kɨt-} to \gl{1+2}\gl{s_p_} (completed in \PWai, ongoing in \kaxui)
	\end{enumerate}
	\item \PWai \begin{enumerate}
		\item \gl{1}\gl{s_p_} \rc{k-} to \gl{1}\gl{s_a_}
		\item innovative \rc{owɨro j-} \qu{\gl{1}\gl{pro} \gl{lk}} for \gl{1}\gl{p}
	\end{enumerate}
	\item \waiwai \begin{enumerate}
		\item \gl{2}\gl{s_a_} \obj{m-} to \gl{2}\gl{s_p_}\end{enumerate}
\end{enumerate}
%
All innovations are person marker extensions except 2b, which combined a pronoun with the linker \rc{j-}.
They are printed in bold in \cref{fig:par_ext}, which reproduces \posscite{gildea1998} tables as a tree diagram, with adapted transcription and an additional \kaxui \gl{1}\gl{s_p_} marker ∅/\obj{j-} \perscommpar{Spike Gildea}.
%
\begin{figure}[hbt]
	\centering
	\setlength{\tabcolsep}{2pt}
	\fbox{\begin{tikzpicture}[
			every node/.append style={align=center},
			level distance=80pt,
			]
			\Tree[.{\PC{}\pcone}
			[.{\PPar{}\ppone}
			{\kaxui{}\kaxone}
			[.{\PWai{}\pwone}
			{\hixka{}\hixone}
			{\waiwai{}\waione}
			]
			]
			]
	\end{tikzpicture}}
	\caption{Person marking extensions in Parukotoan, after \textcite[94]{gildea1998}}
	\label{fig:par_ext}
\end{figure}
%
\hixka has preserved split-\gl{s} only in the second person prefixes, while \kaxui still shows the variation in the first person and \gl{1+2} prefixes that is reconstructible to \PPar.
\waiwai, on the other hand, has lost the system entirely, which notably happened via distinct innovations at three different diachronic stages.
%In this case, the loss of the inflectional classes also entailed the loss of the other morphological traces of the split-\gl{s} system.
%That is, the \gl{s_a_} class marker \rc{w-} \pcref{sec:split} was lost in \waiwai, as evidenced by the contrast between the \waiwai and \kaxui deverbal forms in \exref{wailoss.wai-158} and \exref{wailoss.kax-138}.
%Similarly, the \gl{2}\gl{s_p_} prefix \obj{a-} was extended to imperatives of (former) \gl{s_a_} verbs \exref[wailoss.wai-112]{wailoss.hix-43}, but only C-initial ones \parencite[62]{waiwaihawkins1998}.

\textcite{gildea1998} discusses person marker extensions in the context of the loss of the split-\gl{s} system and the accompanying changes to indexing alignment; for our story of conservative verbs we will zoom in on a particular aspect of these extensions.
To begin, we argue that they took place via lexical diffusion, characterized as a type of extension by \textcite[106--115]{harris1995historical}; this hypothesis is supported by three facts.
First, the variation in first person and \gl{1+2} prefixes described above for \kaxui is not completely free.
Rather, some verbs only allow for example first person \obj{k-}, but not \obj{j-}, while others can occur with both, which is the expected pattern in a lexical diffusion scenario.
In addition, this is speaker-dependent \perscommpar{Spike Gildea}, which is what one would expect from a change in progress.
%todo maybe put in GOOD kaxui data
Second, while there is no detailed diachronic scenario for the switch of \gl{1}>\gl{3} \rc{t-} and \gl{1}\gl{s_a_} in the Tiriyoan languages \pcref{sec:taranoan}, \textcite[111--112]{meira1998proto} convincingly argues that it must have happened gradually rather than instantaneously, and entailed both markers spreading at the same time.
Whether this gradual switch was along ordered lines or not, lexical diffusion must have played a role.

Our third argument in favor of the lexical diffusion scenario brings us back to the conservative \hixka and \trio forms in \cref{tab:hixintro,tab:triintro}.
In both cases, the innovative \gl{1}\gl{s_a_} prefixes were introduced by a person marker extension spreading via lexical diffusion.
We interpret the continued presence of the old \gl{1}\gl{s_a_} prefix in a few verbs as the extension stopping short of these verbs, rather than affecting all targets (\gl{s_a_} verbs).
In our family-wide investigation of person marker extensions, we identified 19 distinct extensions affecting intransitive verbs, and found 6 of them to be incomplete.
These incomplete extensions have left between 1 and 7 conservatively inflected verbs in 9 Cariban languages. %, marked with bold print in \cref{fig:caribantree}.

Interestingly, all six extensions featured innovative first person markers on \gl{s_a_} verbs.
While extensions occurred with other person values as well, they never affected \gl{s_a_} verbs, only \gl{s_p_} ones, and they always affected all potential targets.
Illustrative examples for complete extensions are shown in \cref{tab:completed}: the extension of \gl{1+2}\gl{s_a_} \obj{s(ɨ)-} (< \rc{kɨt-}) to \gl{s_p_} verbs in \apalai \pcref{tab:apalai}, of \gl{2}\gl{s_a_} \obj{m(ɨ)-} in to \gl{s_p_} verbs in \panare \pcref{tab:panare},\footnote{The presence of the third person marker \obj{n-} for \gl{1+2} is due to the wholesale loss of that inflectional value.} or the extension of the entire \gl{s_a_} set to \gl{s_p_} verbs in \waimiri \pcref{tab:waimiri}. %, as well as the extension of \rc{k-} to \gl{1}\gl{s_p_} and of \rc{tɨt-} to \gl{1+2}\gl{s_p_} in \PWai \pcref{fig:par_ext}.
The markedly different behavior of \gl{s_a_} and \gl{s_p_} verbs with regards to the extent of extensions affecting them points to the split-\gl{s} system playing some role, so we will discuss its main properties in \cref{sec:split}.
This will also give readers an idea how it is possible for the \gl{s_a_}/\gl{s_p_} distinction to be lost only for a single person, or for \gl{s_p_} verbs to take on \gl{s_a_} markers with apparent semantic impunity.

\input{floats/completed}

\subsection{The Cariban split-\gl{s} system}
\label{sec:split}
As seen in \cref{sec:pc_person}, the \PC distinction between \gl{s_a_} and \gl{s_p_} verbs was instantiated by two inflection classes within the \setone person paradigm, but this was not the only inflectional difference:
%\begin{inlinelist}
%	\item presence vs absence of the \gl{s_a_} marker \rc{w-}
%	\item absence vs presence of the second person prefix \rc{ə(j)-} in imperatives
%	\item presence vs absence of a derivational detransitivizing prefix
%\end{inlinelist}.
%
Many languages show an \gl{s_a_} class marker in deverbalized forms, originating in \PC \rc{w-}.\footnote{See \textcite[227]{meira2000split}, who identifies reflexes of this morpheme as having \dbqu{no purpose other than being \qu{class markers}, without any obvious semantic or functional load}.}
With \gl{s_a_} verbs, \rc{w-} occurred immediately between the possessive prefixes and the verb stem, while \gl{s_p_} verbs took the bare prefixes.
Reflexes of \rc{w-} in languages from different branches are illustrated in \cref{tab:participles,tab:nominalizations} for participles and nominalizations.%
%
\input{floats/participles}%
%
\input{floats/nominalizations}%
%
\input{floats/imperatives}
%
The distinction between \gl{s_a_} and \gl{s_p_} was also borne out in imperatives, where the latter took the \gl{p}-oriented second person prefix \rc{ə(j)-}, while the former received no prefix (both suffixed with \rc{-kə}).
This is illustrated with reflexes in various modern languages in \cref{tab:imperatives}.
Both the \gl{s_a_} marker \rc{w-} and the prefix pattern in imperatives have been lost in some languages, but are reconstructible to \PC.
%However, there was one further property uniting most \gl{s_a_} verbs, not based on inflectional morphology.

In the modern languages maintaining the split-\gl{s} system, mismatches between the semantics of verbs and their \gl{a}- or \gl{p}-oriented inflectional morphology are common, exemplified with \kalina data in \exref[kar1]{kar2}.

\pex<kar1> \kalina
\a<kar-60>
\begingl
\gla mi-kupi-ja//
\glb \gl{2}>\gl{3}-bathe-\gl{prs}//
\glft \qu{You bathe him/her.} \parencite[][160]{hoff1968carib}//
\endgl
\a<kar-62>
\begingl
\gla a-kupi-ja//
\glb \gl{3}>\gl{2}-bathe-\gl{prs}//
\glft \qu{S/he bathes you.} \parencite[][63]{yamada2011evidentiality}//
\endgl
\xe
%
In \exref{kar1}, the choice between the second person \gl{a}- and \gl{p}-oriented markers \obj{mi-} and \obj{a-} depends on the scenario:
The transitive verb \obj{kupi} \qu{to bathe} takes \obj{mi-} in \gl{2}>\gl{3} scenarios \exref{kar1.kar-60}, but \obj{a-} in \gl{3}>\gl{2} scenarios \exref{kar1.kar-62}.
The intransitive verbs in \exref{kar2} show the same person markers, but the choice of marker depends on the verb, rather than semantics.

\pex<kar2> \kalina
\a<kar-59>
\begingl
\gla sipi tɨnka-rɨ m-ekema-non hen//
\glb net pull-\gl{nmlz} \gl{2}-be.afraid-\gl{prs}.\gl{uncert} eh?//
\glft \qu{You're afraid to pull up the net, aren't you? } \parencite[][253]{courtz2008carib}//
\endgl
\a<kar-61>
\begingl
\glpreamble aya:woiya //
\gla aj-awomɨ-ja//
\glb \gl{2}-get.up-\gl{prs}//
\glft \qu{You are getting up.} \parencite[][167]{hoff1968carib}//
\endgl
\xe
%
Thus, \obj{ekema} \qu{to be afraid} takes an \gl{a}-oriented marker, since it is an \gl{s_a_} verb \exref{kar2.kar-59}, while the \gl{s_p_} verb \obj{awomɨ} \qu{to get up} takes a \gl{p}-oriented marker \exref{kar2.kar-61}.
In both cases, the prefix does not appear to contribute to the semantics of the predicate, since there are clear mismatches:
\qu{to be afraid} with an \dbqu{agentive} marker can hardly be considered a volitional act, while  \qu{to get up} with a \dbqu{patientive} marker must be considered volitional.
\textcite{meira2000split} takes a sizable corpus of intransitive verbs from \trio, \kalina, \apalai, and \wayana, and categorizes them by applying different criteria commonly encountered in split-\gl{s} systems.
He shows that neither (non\-)activities, % \parencite[210]{meira2000split}, 
(non-)agency, % \citeyearpar[212]{meira2000split}, 
(in-)animacy, % \citeyearpar[213]{meira2000split},
nor Aktionsart % \citeyearpar[215]{meira2000split} 
satisfactorily predict the class membership of intransitive verbs in any of the languages.

Rather, the reason for a verb to take \gl{a}- or \gl{p}-oriented prefixes is (at least diachronically) a morphological one.
\textcite[217--221]{meira2000split} demonstrates that those intransitive verbs which (etymologically) have a detransitivizing prefix are treated as \gl{s_a_} verbs, while essentially all others are \gl{s_p_} verbs:
\begin{quotebox}{\parencite[201]{meira2000split}}
	Almost all verbs in the \gl{s_a_} class are detransitivized forms of transitive verbs, either synchronically (with still exisiting transitive sources) or diachronically (with reconstructible but no longer existing transitive sources)
\end{quotebox}
\textcite[221--223]{meira2000split} also argues that the detransitivizing prefixes are indeed deriving \gl{s_a_} verbs, rather than being inflectional in nature:
\begin{inlinelist}
	\item there are a few underived \gl{s_a_} verbs, with no detransitivizing prefix
	\item \gl{s_a_} verbs can develop irregular semantics compared to their transitive counterparts
	\item it is unpredictable whether the \gl{a} or \gl{p} argument of the underlying transitive verb becomes the \gl{s} of the derived \gl{s_a_} verb
	\item some originally derived \gl{s_a_} verbs have lost their transitive counterparts
	\item \dbqu{basic} concepts are expressed as derivations of more complex concepts, like \qu{to dance (\gl{s_a_})} from \qu{to dance with (\gl{tr})}
\end{inlinelist}.
He also notes that this leads to an inflectional split not based in meaning, but rather morphology:

@@

As for the form of this marker, \textcite[505--512]{meira2010origin} reconstruct two distinct  prefixes for \PC: reciprocal \rc{əte-} and reflexive \rc{e-}, although on verbs their reflexes have been merged in a single morpheme in modern languages.
Modern reflexes of \detrz show a range of meanings, which can all be characterized as \dbqu{detransitive}; this range is illustrated with \trio examples in \exref{triodetrz}.

\ex<triodetrz> \trio \parencites[218--219]{meira2000split}[128, 256]{triomeira1999}\\
\begin{tabular}[t]{@{}llll@{}}
	\\
	\obj{nonta}  & → & \obj{e-nonta}, & \qu{abandon each other}\\
	\qu{abandon} & & \obj{əi-nonta} &  (reciprocal) \\
	\\
	\obj{suka} & → & \obj{e-suka}, & \qu{wash self}\\
	\qu{wash} & & \obj{əi-suka} & (reflexive)\\
	\\
	\obj{pahka} & → & \obj{e-pahka} & \qu{break (\gl{intr})}\\
	\qu{break (\gl{tr})} & & & (anticausative)\\
	\\
	\obj{puunəpɨ} & → & \obj{əh-puunəpɨ}, & \qu{think, meditate}\\
	\qu{think about} & & \obj{əi-puunəpɨ} & (antipassive)\\
\end{tabular}
\xe
%
The morphological variation featured in \qu{to abandon each other} and \qu{to wash self} is due to the mentioned collapse between the two \PC prefixes:
\obj{e-} is a reflex of the reflexive prefix \rc{e-}, while the form \obj{əi-} originates in reciprocal \rc{əte-}.
However, both can occur with either meaning -- at least for these two verbs.
%As mentioned, there are a few \gl{s_a_} verbs that are not derived from transitive verbs, and do not have a reflex of \detrz, either.
%\textcite[221]{meira2000split} characterizes these as very old and exceptional, and counts between 5 and 7 verbs, depending on the language.
%\textcite{gildea2007greenberg} identify 7 such underived \gl{s_a_} verbs as reconstructible to \PC, shown in \cref{tab:underived2007}.
%Many of these verbs will be discussed in \cref{sec:verbs}.

%\input{floats/underived2007.tex}
%
%
%\subsection{Summary and outlook}
%\label{sec:outlook}
%\gl{s_a_} verbs irregularly inflected for first person were identified in 9 Cariban languages.
%These irregular forms are actually conservative, belonging to a few verbs unaffected by person marker extensions spread via lexical diffusion.
%Such conservative forms are only found among the first person forms of  (etymological) \gl{s_a_} verbs.
